---
title: "Data in R."
author: "Katlyn H. Degamo"
date: "2023-05-09"
output: pdf_document
---

```{r}

 #To store something in R session, you will assign it a name using <- operator. 
 #You can assign your previous calculation to an object called my_sum.
 #Simple calculation like this.

3 + 3
my_sum <- 4 + 4
my_sum
my_sum + 5

 #To display the value, the name of the object must be typed, the *print()* command or 
 #used the command should be wrapped in parenthesis.

my_sum
print(my_sum)
(another_sum <- 7 + 8)
(new_sum <- my_sum + 5)
```

## Data Types

```{r}
 #The typeof function can be used to see the type of a single scalar value.
 #integer can be signified by adding an 'L' to the end.
 #Numeric data can be integer form or double form.

my_integer <- 2L
typeof(my_integer)
my_double <- 5.94
typeof(my_double)
 
 #Character data is text data sorrounded by single or doule qoutes.
my_character <- "HI I'M THE PROBLEM"
typeof(my_character)
 
 #Logical data takes the form TRUE or FALSE.

my_logical <- FALSE
typeof(my_logical)
```

## Homogeneous Data Structures

```{r}
 #Vectors are one-dimensional structures containing data of the same type, 
 #notated by using c().
 #Type of vector can also be viewed using the typeof function, 
 #but str function can be used to display both contents and its type.

double_vector <- c(1.2, 2.3, 3.4, 45, 56, 7)
typeof(double_vector)
str(double_vector)
 
 #Categorical data which takes only a finite number of possible values.

categories <- factor(c("A", "B", "C", "A", "A", "C"))
str(categories)
categories_char <- c("A", "B", "C", "A", "A", "C")
str(categories)

 #If needed, the factors can be given order.
 #Character vector.

ranking <- c("Medium", "High", "Low")
str(ranking)

 #turn it into ordered factor.

ranking_factors <- ordered(ranking, levels = c("Low", "Medium", "High"))
str(ranking_factors)

 #number of elements in a vector can be seen using the length() function.

length(categories)
categories
length(ranking_factors)

 #simple numeric sequence vectors can be created using shorthand notation.

(my_sequence <- 1:10)
seq(1, 10, 2)
seq(2, 20, 2)
seq(from = 1, to = 20, by = 5)

 #numeric sequence vector.
 #create a new vector containing vec and the character "HELLO".
 #numeric values have been coerced into their character equivalents.

vec <- 1:10
str(vec)
new_vec <- c(vec, "HELLO")
str(new_vec)

 #attempt a mixed logical and numeric.
 #logical has been converted to binary numeric (TRUE = 1)
 #try to add a numeric to our previous categories factor vector.
 #categories have been coerced to background integer representations.

vec[1] + vec [2]
mix <- c(TRUE, 6)
str(mix)
new_categories <- c(categories,1)
str(new_categories)
str(categories)

 #Matrices are two-dimensional data structures of the same type and 
 #are built from a vector by defining the number of rows and columns.

(m <- matrix(c(1, 2, 3, 4), nrow = 2, ncol = 2))
(m <- matrix(vec, nrow = 5, ncol = 2))
```

## Heterogeneous Data Structure

```{r}
 #List are one-dimensional data structures that can take data of any type.

my_list <- list(6, TRUE, "HELLO")
str(my_list)
new_list <- list(scalar = 6, some_vec = c("HELLO"), some_mat = matrix(1:4, nrow = 2, ncol = 4))
str(new_list)
new_list[[3]]
new_list[[2]]
new_list[[1]]

 #named list elements can be accessed by using $.

new_list$scalar
new_list$some_vec
new_list$some_mat

 #Data frames - are the most used data structure in R; 
          #very similar in nature to a typical database table or spreadsheet.
          #two vectors of different types but same length.

names <- c("John", "Ayesha")
ages <- c(31, 24)

 #Create a data frame.

(df <- data.frame(names, ages))

 #Get types of columns

str(df)

 #Get dimensions of df

dim(df)
```

# Working with Dataframes

```{r}
 #Dataframes is the most common data structure used by analysts in R.
 #R facilitates numerous ways of importing data from simple .csv files, from Excel files
  #from online sources or from databases.
```

## Loading and tidying data in dataframes

```{r} 
 #Let's load the salespeople data set, which contains some information on the sales,
  #average customer ratings and performance ratings of salespeople.
 #read.csv() function can accept a URL address of the file if it is online.

 #url of data set
 #dataframe called salespeople

data()
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)

 #View the dimensions.

dim(salespeople)

 #If it is too big to display, we can use head() function.

head(salespeople)

 #If we wanted to see the 6th entry of sales column.
 #We can view a specific column by using $, and we can use brackets to view specific entry.

salespeople$sales[6]
salespeople[34, 4]
salespeople

 #We can take a look at the data types using str().

str(salespeople)

 #We can also see a statistical summary of each column using summary().
 #Missing data is identified by a special NA value in R.

summary(salespeople)

 #The function is.na() will look at all values in a vector or dataframe 
 #and return TRUE or FALSE 
  #based on whether they are NA or not.

is.na(salespeople)

 #By adding these up using the sum() function, it will take TRUE as 1 and FALSE as 0,
  #which effectively provides a count of missing data.

sum(is.na(salespeople))

 #The easiest way is to use the complete.cases() function, 
 #which identifies the rows that have no NAs, 
  #and then we can select those rows from the dataframe based on that conditon.

complete.cases(salespeople)
salespeople[complete.cases(salespeople),]
salespeople <- salespeople[complete.cases(salespeople),]

 #confirm no NAs.

sum(is.na(salespeople))

 #We can see the unique values of a vector or column using the unique() function.

salespeople$performance
unique(salespeople$performance)

 #If we need to change the type of a column in a dataframe, we can use the 
  # as.numeric(), as.character(), as.logical() or as.factor() functions.

as.factor(salespeople$performance)
str(salespeople)
salespeople$performance <- as.factor(salespeople$performance)
str(salespeople)
```

## Manipulating dataframes

```{r}
 #Dataframes can be subsetted to contain only rows that satisfy specific conditions.
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)

(sales_720 <- subset(salespeople, subset = sales == 720))

 #Note the use of ==, which is used in many programming languages, 
  #to test for precise equality.

unique((salespeople$sales))

 #Similarly we can select columns based on inequalities

(subset(salespeople, subset = sales >=700))
high_sales <- subset(salespeople, subset = sales >= 700)
head(high_sales)

 #To select specific columns use the select argument.

salespeople_sales_perf <- subset(salespeople, select = c("sales", "performance"))
head(salespeople_sales_perf)

 #bind the rows of low_sales and high_sales together.

low_sales <- subset(salespeople, subet = sales < 400)
low_and_high_sales = rbind(low_sales, high_sales)
head(low_and_high_sales)
head(high_sales)

 #Two dataframes with two columns each.
 #Bind the columns to create a dataframe with four columns.

sales_perf <- subset(salespeople, select = c("sales", "performance"))
prom_custrate <- subset(salespeople, select = c("promoted", "customer_rate"))
full_df <- cbind(sales_perf, prom_custrate)
head(full_df)

 #In the code so far we have used a variety of functions.
 #Exampe head(), subset(), rbind().
 
 #Functions exist to perform common useful operations.
 
 #Often there are a large number of arguments that a function can take,
  #but many are optional and not required to be specified by the user.
 #Example, the function head(), which displays the first rows of a dataframe
  #has only on required argument x: the name of the dataframe.

head(prom_custrate)
head(full_df, n = 5)

head(x = salespeople)
head(x = salespeople, n=4)

 #When running a function, you can either specify the arguments by name or 
  #you can enter them in order without their names.

 #see fewer rows - arguments to be in the right order if not named.

head(salespeople, 3)
head(salespeople, 6)

 #or if you don't know the right order name your arguments 
 #and you can put them in any order.

head(n=3, x = salespeople)
head(n=6, x = salespeople)
```

# Functions, packages and libraries

```{r}
 #Most functions in R have excellent help documentations
 #To get help on the head() function, type help(head) or ?head.

?head
help(head)
 
 #Alternatively you can open the Help browser window directly in RStudio 
 #and do a search there.
```

## Writing your own functions

```{r}
 #Functions are not limited to those that come packaged in R.
 #Users can write their own functions to perform tasks 
 #that are helpful to their objectives.

 #In this example, a simple function is written 
 #which generates a report on a dataframe

df_report <- function(df){paste("This dataframe contains", nrow(df), "rows and", ncol(df), "columns. 
                                There are", sum(is.na(df)), "NA entries.")}

 #We can test our function by using the built-in mtcars data set in R.

df_report(mtcars)
paste("This is how the paste", "functioning works")
df_report(iris)
```

## Installing packages

```{r setup, include=FALSE}
 #A package is an additional module that can be installed easily.
 #we will be using functions from both base R and from popular and useful packages.
 #a popular package used for statistical modeling is the MASS package, 
 #which  is based on methods in a popular applied statistics book.

#install.packages("MASS")
library(MASS)
 #a popular package used for statistical modeling is the MASS package, 
 #which is based on methods in a popular applied statistics book.

my_packages <- c("DescTools", "dplyr")
#install.packages(my_packages)
library(DescTools, dplyr)
```

##  Using packages

```{r}
 #Once you have installed a package into your package library, 
 #to use it in your R session you need to load it using the library() function

library(MASS)
help(package = "MASS")

 #Once a package is loaded from your library, 
 #you can use any of the functions inside it.

#MASS::stepAIC()
```
### The pipe operator

```{r}
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)

 #The pipe operator makes code more natural to read and write and reduces 
 #the typical computing problem of many nested operations inside parentheses.

sample(14,14)
sales <- subset(salespeople , subset = sales < 500)

 #take the mean of those values

mean(sales$sales)
mean(subset(salespeople$sales, subset = salespeople$sales < 500))
mean(subset(salespeople, subset = sales < 500)$sales)

 #load magrittr library to get the pipe operator

library(magrittr)
subset(salespeople$sales, subset = salespeople$sales < 500) %>% 
mean() 
round(mean(subset(salespeople, subset = sales < 500)$sales))
```

# Errors, warnings and messages

```{r}
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)

 #Errors are serious problems which usually result in the halting of your code 
 #and a failure to return your requested output.

 #subset(salespeople, subset = sales = 720) - the equal sign should be double

subset(salespeople, subset = sales == 720)  

 #head[salespeople] - parentheses, not brackets

head(salespeople)

 #Warnings are less serious and usually alert you to something that 
 #you might be overlooking and which could indicate a problem with the output.
 #Messages are pieces of information that may 
 #or may not be useful to you ata particular point in time. 
```

#  Plotting and graphing

```{r}
 #there are numerous ways to plot and graph information in R. 
 #graphical output will be rendered in the Plots pane, 
 #where you can copy it  or save it as an image.
```

##  Plotting in base R

```{r}
url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)

 #The simplest plot function in base R is plot(). This performs basic X-Y plotting.
 #Note the use of the arguments main, xlab and ylab 
 #for customizing the axis labels and title for the plot.

plot(x = salespeople$sales, y = salespeople$customer_rate)

plot(x = salespeople$sales, y = salespeople$customer_rate,
     xlab = "Sales ($m)", ylab = "Average customer rating",
     main = "Scatterplot of Sales vs Customer Rating")

  #Histograms of data can be generated using the hist() function.
  #Note the use of breaks to customize how the bars appear.

hist(salespeople$performance)
hist(salespeople$performance,
     xlab = "Performance Rating", 
     main = "Histogram of Performance Ratings")

hist(salespeople$performance, breaks = 0:4,
     xlab = "Performance Rating", 
     main = "Histogram of Performance Ratings")

 #Box and whisker plots are excellent ways to see the distribution of a variable, 
 #and can be grouped against another variable to see bivariate       patterns.
 #Note the use of the formula and data notation here to define the variable 
 #we are interested in and how we want it grouped. 

boxplot(formula = sales ~ performance, data = salespeople)

boxplot(formula = sales ~ performance, data = salespeople,
        xlab = "Performance Rating", ylab = "Sales ($m)",
        main = "Boxplot of Sales by Performance Rating")

```

## Specialist plotting and graphing packages

```{r}
 #the most commonly used specialist plotting and graphing package in R is ggplot2.
 #ggplot2 allows the flexible construction of a very wide range of charts and  graphs, 
 #but uses a very specific command grammar which can take some getting  used to.

#install.packages("ggplot2")
library(ggplot2)
 #The plotly package allows the use of the plotly graphing library in R. 
 #This is an excellent package for interactive graphing 
 #and is used for 3D illustrations in this book.

#install.packages("plotly")

 #GGally is a package that extends ggplot2 to allow easy combination of charts and graphs. 
 #One of its most popular functions is ggpairs(), which produces a pairplot. 

install.packages("GGally")

url <- "https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv"
salespeople <- read.csv(url)
salespeople$promoted <- as.factor(salespeople$promoted)
salespeople$performance <- as.factor(salespeople$performance)
GGally::ggpairs(salespeople)
```

# References

- [ Venables, W. N., and B. D. Ripley. 2002. Modern Applied Statistics with s.Wickham, Hadley. 2016. ggplot2: Elegant Graphics for Data Analysis.](https://ggplot2-book.org/)
- [Wickham, Hadley, and Garrett Grolemund. 2016. R for Data Science.]( https://r4ds.had.co.nz/)
- [Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. R Markdown Cookbook.](https://bookdown.org/yihui/rmarkdown-cookbook/)
- [salespeople](https://raw.githubusercontent.com/msuiitdmsgabriel/datasets-regression/main/salespeople.csv)
 
