---
title: "Simple Linear Regression - Exam"
author: "Katlyn H. Degamo"
date: "2023-06-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. The fractional Distillation Data

```{r}
x_hydrocarbon <- c(1.02, 1.11, 1.43, 1.11, 1.01, 0.95, 1.11, 0.87, 1.43, 1.02,
                 1.46, 1.55, 1.55, 1.55, 1.40, 1.15, 1.01, 0.99, 0.95, 0.98)
y_purity <- c(86.91, 89.85, 90.28, 86.34, 92.58, 87.33, 86.29, 91.86, 95.61, 89.86,

                    96.73, 99.42, 98.66, 96.07, 93.65, 87.31, 95.00, 96.85, 85.20, 90.56)
x = x_hydrocarbon 
y = y_purity

```

## a. create a scatter diagram for the table

```{r}
library(ggplot2)

# Display Scatter Diagram 

scatter_diagram <- ggplot(data = NULL, 
                          aes(x = x_hydrocarbon, y= y_purity)) + 
                          geom_point() + geom_smooth(method = "lm", 
                          se = FALSE) + 
                          labs(title = "The Fractional Distillation Data",
                          x = " Hydrocarbon", y = "Purity")

print(scatter_diagram)
```

## b. The least-square fit is

```{r}
# Creating the linear regression model
model <- lm(y ~ x)

# Getting the coefficients
(coefficients <- model$coefficients)

# Getting the R-squared value
(r_squared <- summary(model)$r.squared)

# Getting the p-values for the coefficients
p_values <- summary(model)$coefficients[, "Pr(>|t|)"]
print(p_values)

model_summary <- summary(model)
model_summary

# Function for simple linear regression
simple_linear_regression <- function(x, y) {
  # Creating the linear regression model
  model <- lm(y ~ x)
  
  # Extracting coefficients
  coefficients <- model$coefficients
  
  # Extracting R-squared value
  r_squared <- summary(model)$r.squared
  
  # Returning coefficients and R-squared value
  return(list(coefficients = coefficients, r_squared = r_squared))
}

result <- simple_linear_regression(x, y)
print(result$coefficients)
print(result$r_squared)



#Creating Linear Model
model_names <- names(model)

# Inspecting the names of objects in the model
print(model_names)
```

```{r}
# Function to get the least-squares fit coefficients
leastsquaresfit <- function(model){
  leastsfit <- model$coefficients
  return(leastsfit)
}

leastfits <- leastsquaresfit(model)

output <- paste("The least-squares fit coefficients:",
                paste("Intercept =", leastfits[1]),
                paste("PropellantAge =", leastfits[2]), sep = "\n")
message(output)
```

```{r}
least_fits <- lsfit(x, y)$coefficients
print(least_fits)
```

## c. The estimate of sigma squared is

```{r}
getsigmasquared <- function(model){
  modelsummary <- summary(model)
  sigmasquared <- modelsummary$sigma^2
  return(sigmasquared)
}

estimatesigmasquared <- getsigmasquared(model)

output <- paste("The Estimate of Sigma Squared is equal to:", estimatesigmasquared)

message(output)
```

## d. Test for significance of regression in the regression model.

```{r}
# Another way to find the p value
p_value <- summary(model)$coefficients[2, 4]

# Print the p-value
print(p_value)
```

```{r}
# Get the coefficient and standard error for a specific variable, e.g., "x"
coefficient <- coef(model)["x"]
standard_error <- summary(model)$coefficients["x", "Std. Error"]

# Calculate the t-value
t_value <- coefficient / standard_error

# Print the t-value
print(t_value)
```

```{r}
# Set the significance level
significance_level <- 0.05

# Set the degrees of freedom
degrees_of_freedom <- 18

# Calculate the critical value
critical_value <- qt(1 - (significance_level / 2), df = degrees_of_freedom)

# Print the critical value
print(critical_value)
```

```{r}
# Perform the significance test for regression
if (abs(t_value) > critical_value) {
    null_hypothesis_status <- "The null hypothesis is rejected."
  } else {
    null_hypothesis_status <- "The null hypothesis is not rejected."
  }

output <- paste("Significance test for regression:",
                paste("T Value =", t_value),
                paste("Critical Value =", critical_value),
                paste(null_hypothesis_status), sep = "\n")

message(output)
```

## e. Use an analysis-of-variance approach to test significance of regression.

```{r}
# Set the significance level
significance_level <- 0.01

# Specify the degrees of freedom for the numerator and denominator
df_numerator <- 1  # Degrees of freedom for the numerator
df_denominator <- 18  # Degrees of freedom for the denominator

# Calculate the critical value
critical_value <- qf(1 - significance_level, df_numerator, df_denominator)

# Print the critical value
print(critical_value)
```

```{r}
# Perform the analysis-of-variance approach to test significance of regression
if (model_summary$fstatistic["value"] > critical_value) {
  null_hypothesis_status <- "The null hypothesis is rejected."
}else { 
  null_hypothesis_status <- "The null hypothesis is not rejected."
}

output = paste("The Analysis Of Variance:",
              paste("F Statistics:", model_summary$fstatistic["value"] ),
              paste("Critical Value:", critical_value),
              paste(null_hypothesis_status), sep = "\n")

message(output)
```

## f. Find a 95%CI on the slope

```{r}
# Calculate the confidence interval 
conf_interval <- confint(model, level = 0.95)

conf_interval
```

## g. Find a 95%CI on the mean purity when the hydrocarbon percentage is 1.00

```{r}
x_pred <- 1.00
y_pred <- predict(model, newdata = data.frame(x =x_pred), interval= "confidence", level = 0.95)
y_ci <- y_pred[, c("lwr", "upr")]
y_ci
```


# 2. The Steam Consumption Data

```{r}
x_tempearture <- c(21, 24, 32, 47, 50, 59, 68, 74, 62, 50, 41, 30)
y_usage <- c(185.79, 214.47, 288.03, 424.84, 454.68, 539.03, 621.55, 675.06, 562.03, 452.93, 369.95, 273.98)
x = x_tempearture
y = y_usage
```

## a. create a scatter diagram for the table

```{r}
library(ggplot2)

# Display Scatter Diagram 

scatter_diagram <- ggplot(data = NULL, 
                          aes(x = x_tempearture, y= y_usage)) + 
                          geom_point() + geom_smooth(method = "lm", 
                          se = FALSE) + 
                          labs(title = "The Steam Consumption Data",
                          x = " Tempearture", y = "Usage")

print(scatter_diagram)
```

## b. The least-square fit is

```{r}
# Creating the linear regression model
model <- lm(y ~ x)

# Getting the coefficients
(coefficients <- model$coefficients)

# Getting the R-squared value
(r_squared <- summary(model)$r.squared)

# Getting the p-values for the coefficients
p_values <- summary(model)$coefficients[, "Pr(>|t|)"]
print(p_values)

model_summary <- summary(model)
model_summary

# Function for simple linear regression
simple_linear_regression <- function(x, y) {
  # Creating the linear regression model
  model <- lm(y ~ x)
  
  # Extracting coefficients
  coefficients <- model$coefficients
  
  # Extracting R-squared value
  r_squared <- summary(model)$r.squared
  
  # Returning coefficients and R-squared value
  return(list(coefficients = coefficients, r_squared = r_squared))
}

result <- simple_linear_regression(x, y)
print(result$coefficients)
print(result$r_squared)



#Creating Linear Model
model_names <- names(model)

# Inspecting the names of objects in the model
print(model_names)
```

```{r}
# Function to get the least-squares fit coefficients
leastsquaresfit <- function(model){
  leastsfit <- model$coefficients
  return(leastsfit)
}

leastfits <- leastsquaresfit(model)

output <- paste("The least-squares fit coefficients:",
                paste("Intercept =", leastfits[1]),
                paste("PropellantAge =", leastfits[2]), sep = "\n")
message(output)
```

```{r}
least_fits <- lsfit(x, y)$coefficients
print(least_fits)
```

## c. The estimate of sigma squared is

```{r}
getsigmasquared <- function(model){
  modelsummary <- summary(model)
  sigmasquared <- modelsummary$sigma^2
  return(sigmasquared)
}

estimatesigmasquared <- getsigmasquared(model)

output <- paste("The Estimate of Sigma Squared is equal to:", estimatesigmasquared)

message(output)
```

## d. Test for significance of regression in the regression model.

```{r}
# Another way to find the p value
p_value <- summary(model)$coefficients[2, 4]

# Print the p-value
print(p_value)
```

```{r}
# Get the coefficient and standard error for a specific variable, e.g., "x"
coefficient <- coef(model)["x"]
standard_error <- summary(model)$coefficients["x", "Std. Error"]

# Calculate the t-value
t_value <- coefficient / standard_error

# Print the t-value
print(t_value)
```

```{r}
# Set the significance level
significance_level <- 0.05

# Set the degrees of freedom
degrees_of_freedom <- 10

# Calculate the critical value
critical_value <- qt(1 - (significance_level / 2), df = degrees_of_freedom)

# Print the critical value
print(critical_value)
```

```{r}
# Perform the significance test for regression
if (abs(t_value) > critical_value) {
    null_hypothesis_status <- "The null hypothesis is rejected."
  } else {
    null_hypothesis_status <- "The null hypothesis is not rejected."
  }

output <- paste("Significance test for regression:",
                paste("T Value =", t_value),
                paste("Critical Value =", critical_value),
                paste(null_hypothesis_status), sep = "\n")

message(output)
```

## e. Use an analysis-of-variance approach to test significance of regression.

```{r}
# Set the significance level
significance_level <- 0.01

# Specify the degrees of freedom for the numerator and denominator
df_numerator <- 1  # Degrees of freedom for the numerator
df_denominator <- 10  # Degrees of freedom for the denominator

# Calculate the critical value
critical_value <- qf(1 - significance_level, df_numerator, df_denominator)

# Print the critical value
print(critical_value)
```

## f. Find a 95%CI on the slope

```{r}
# Calculate the confidence interval 
conf_interval <- confint(model, level = 0.95)

conf_interval
```

## g. Find a 95%CI on the mean purity when the hydrocarbon percentage is 1.00

```{r}
x_pred <- 1.00
y_pred <- predict(model, newdata = data.frame(x =x_pred), interval= "confidence", level = 0.95)
y_ci <- y_pred[, c("lwr", "upr")]
y_ci
```


